Supplementary Materials

### 2023-02-07 analysis/paper/paper.qmd
rename analysis directory

library(here)
library(readr)

# upload data
metadata <- read_tsv(here("inst/data/raw_data/metadata.tsv"))
demography <- read_csv(here("inst/data/raw_data/demography.csv"))
lloq <- read_tsv(here("inst/data/raw_data/lloq.tsv"))
uhplc_data_comb <- read_csv(here("inst/data/derived_data/uhplc-data_combined.csv"))
#uhplc_calculus <- read_csv(here("inst/data/derived_data/uhplc-calculus_cleaned.csv"))
dental_inv <- read_csv(here("inst/data/raw_data/dental-inv.csv"))
caries <- read_csv(here("inst/data/raw_data/caries.csv"))
periodont <- read_csv(here("inst/data/raw_data/periodontitis.csv"))
periap <- read_csv(here("inst/data/raw_data/periapical.csv"))
calculus <- read_csv(here("inst/data/raw_data/calculus.csv"))
calculus_full <- read_csv(here("inst/data/raw_data/calculus_full.csv"))
sinusitis_clean <- read_csv(here("inst/data/derived_data/sinusitis_cleaned.csv"))
path_cond_clean <- read_csv(here("inst/data/derived_data/path-conditions_cleaned.csv"))

>>>>

library(here)
library(readr)
devtools::load_all()

# upload data
metadata <- read_tsv(here("inst/data/raw_data/metadata.tsv"))
demography <- read_csv(here("inst/data/raw_data/demography.csv"))
lloq <- read_tsv(here("inst/data/raw_data/lloq.tsv"))
uhplc_data_comb <- read_csv(here("inst/data/derived_data/uhplc-data_combined.csv"))
dental_inv <- read_csv(here("inst/data/raw_data/dental-inv.csv"))
caries <- read_csv(here("inst/data/raw_data/caries.csv"))
periodont <- read_csv(here("inst/data/raw_data/periodontitis.csv"))
periap <- read_csv(here("inst/data/raw_data/periapical.csv"))
calculus <- read_csv(here("inst/data/raw_data/calculus.csv"))
calculus_full <- read_csv(here("inst/data/raw_data/calculus_full.csv"))
sinusitis_clean <- read_csv(here("inst/data/derived_data/sinusitis_cleaned.csv"))
path_cond_clean <- read_csv(here("inst/data/derived_data/path-conditions_cleaned.csv"))

### 2023-02-07 analysis/supplementary-materials/supp-mat.qmd
rename analysis directory

---
title: "Supplementary Materials"
knitr: 
  opts_chunk: 
    warning: false
    message: false
format: 
  html:
    toc: true
---

@@

source(here("inst/scripts/setup-qmd.R"), local = knitr::knit_global())
```

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| sample | sample number for the UHPLC-MS/MS analysis |

@@

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and status |

@@

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and location of caries |

@@

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and periodontitis score (0-3) |

@@

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and periapical lesion |

@@

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and deposit size (0-3) |

@@

## Sample demographics

@@

## Missing data

@@

```{r}
#| label: fig-missing-hex
caries %>%
  mutate(across(!id, ~ is.na(.x))) %>%
  select(!id) %>%
  colSums()

caries %>%
  mutate(across(!id, ~ is.na(.x))) %>% 
  pivot_longer(-id) %>% 
  ggplot(aes(x = name, y = id, fill = value)) +
    geom_tile()

@@

```{r}
#| label: fig-missing-dental
dental_long %>%
  mutate(across(c(caries, periodont, periap, calculus), is.na)) %>%
  select(c(caries, periodont, periap, calculus, id, tooth)) %>%
  #group_by(id, tooth) %>% 
  rowwise() %>% 
  mutate(missing = sum(caries, periodont, periap, calculus)) %>% 
  ggplot(aes(x = tooth, y = id, fill = missing)) +
    geom_tile()
```

@@

@@

@@

```{r}
#| label: fig-detection-preservation
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected
    ) %>% 
  ggplot(aes(x = preservation, y = quant, fill = preservation)) +
    geom_violin() +
    geom_jitter(width = 0.2) +
    #geom_boxplot(alpha = 0.6, width = 0.2) +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
# 
# uhplc_calculus_demography %>%
#   filter(
#     !is.na(preservation),
#     !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected
#     ) %>% 
#   ggplot(aes(x = preservation, y = quant, fill = preservation)) +
#     geom_violin() +
#     geom_jitter(width = 0.2) +
#     #geom_boxplot(alpha = 0.6, width = 0.2) +
#     facet_wrap(~ compound, scales = "free_y") +
#     theme_bw()
```

```{r}
#| label: fig-detection-preservation2
#uhplc_calculus_demography %>%
uhplc_calculus_long %>% 
  filter(!is.na(preservation)) %>% 
  ggplot(aes(x = preservation, y = quant, fill = preservation)) +
    geom_boxplot() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
```

@@

@@

@@

@@

@@

Caries were scored based on location in each individual. Multiple locations on
a single tooth is possible. The size of caries was also
recorded, but not used in further analysis. Large caries that cover multiple
surfaces were recorded as 'crown'. Caries counts were extracted based on the
number of locations recorded.

@@

In the `r caries_ratio_site$n_teeth` that were examined, `r caries_ratio_site$count`
teeth had caries
(`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`).
This frequency has very little meaning, and was further broken down into tooth
class and individuals.

@@

@@

>>>>

---
title: "Supplementary Materials"
execute: 
  echo: false
  warning: false
knitr: 
  opts_chunk: 
    message: false
format: 
  html:
    toc: true
    code-fold: true
---

@@

source(here("inst/scripts/setup-qmd.R"))
```

@@

<!-- id	sample	element	KZ_sample	KZ_element	replicated	batch1_weight	batch2_weight -->

| variable | description |
|---|---|
| id | unique identifier for the individual |
| sample | sample number for the UHPLC-MS/MS analysis |
| element | which tooth (FDI notation) was sampled for UHPLC-MS/MS analysis |
| replicated | Whether the sample was included in the replication batch (TRUE/FALSE) |
| batch1_weight | weight (mg) of sample in batch 1 |
| batch2_weight | weight (mg) of sample in batch 2 |

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| t11..t48 | Tooth number (FDI system) and status |

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| t11..t48 | Tooth number (FDI system) and location of caries |

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| t11..t48 | Tooth number (FDI system) and periodontitis score (0-3) |

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| t11..t48 | Tooth number (FDI system) and periapical lesion |

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| t11..t48 | Tooth number (FDI system) and deposit size (0-3) |

@@

## Sample

### Selection

```{r}
#| label: tbl-sampling
#| tbl-cap: "Table showing which individuals were sampled in this study (element) and which individuals were sampled in the previous study (KZ_element)."
metadata %>% 
  select(id, element, KZ_element) %>% 
  mutate(
    element = case_when(is.na(element) ~ FALSE,
                   TRUE ~ TRUE),
    KZ_element = case_when(is.na(KZ_element) ~ FALSE,
                   TRUE ~ TRUE),
    )
```

### Demographics

@@

### Missing data

@@

```{r}
#| label: fig-missing-caries
caries %>%
  mutate(across(!id, ~ is.na(.x))) %>%
  select(!id) %>%
  colSums()

caries %>%
  mutate(across(!id, ~ is.na(.x))) %>% 
  pivot_longer(-id) %>% 
  ggplot(aes(x = name, y = id, fill = value)) +
    geom_tile()
```

@@

```{r}
#| label: fig-missing-dental
#| fig-cap: "Number of missing values per individual and tooth across all dental scores (caries, periodontitis, periapical lesions, calculus)."
dental_long %>%
  mutate(across(c(caries, periodont, periap, calculus), is.na)) %>%
  select(c(caries, periodont, periap, calculus, id, tooth)) %>%
  #group_by(id, tooth) %>% 
  rowwise() %>% 
  mutate(missing = sum(caries, periodont, periap, calculus)) %>% 
  ggplot(aes(x = tooth, y = id, fill = missing)) +
    geom_tile()
```

@@

```{r}
#| label: setup-uhplc

# presence/absence data frame
uhplc_calculus_bin <- uhplc_calculus_long %>%
  mutate(presence = if_else(quant > 0, 1, quant))

# successfully replicated samples only
uhplc_calculus_replicated <- uhplc_calculus_bin %>%
  mutate(compound = str_remove(compound, "_calc")) %>%
  group_by(id, sample, compound) %>% # combine batches
  summarise(presence = sum(presence)) #%>%
  # filter(presence == 0 | presence == 2) %>%
  # group_by(compound) %>%
  # mutate(presence = if_else(presence == 2, 1, 0)) # convert replications to presence/absence

uhplc_calculus_replicated <- uhplc_calculus_bin %>%
  filter(id %in% filter(metadata, replicated == TRUE)$id) %>%
  #right_join(select(filter(metadata, replicated == TRUE)), by = c("id", "sample"))
  #mutate(compound = str_remove(compound, "_calc")) %>%
  group_by(id, sample, compound) %>% # combine batches
  summarise(presence = sum(presence)) %>%
  filter(presence == 0 | presence == 2) %>% # remove compounds only detected in one batch
  group_by(compound) %>%
  mutate(presence = if_else(presence == 0, 0, 1)) %>%  # convert replications to presence/absence
  ungroup()

uhplc_replicated_wide <- uhplc_calculus_replicated %>%
  mutate(compound = case_when(compound == "nicotine" ~ "tobacco",
                              compound == "cotinine" ~ "tobacco",
                              TRUE ~ compound)) %>%
  group_by(id, sample, compound) %>%
  summarise(presence = sum(presence)) %>% # combine nicotine and cotinine
  #remove_missing() %>%
  mutate(presence = case_when(presence > 0 ~ TRUE,
                              TRUE ~ FALSE)) %>%
  pivot_wider(names_from = "compound", values_from = "presence")
```

The UHPLC-MS/MS method was validated in a separate study on cadavers received
for forensic autopsy and toxicological analysis. Results from dental calculus were
validated against compounds detected in whole blood samples from the same
individuals [@sorensenDrugsCalculus2021].  
In the original method, samples were washed three times to remove residual
substances from the surface of the calculus that originated from oral fluids,
and only extract substances from the calculus. In our samples the washes served
to remove potential contaminants from the burial environment and post-excavation
handling.

Briefly, dental calculus was treated with citric acid and the dissolution extracts
were cleaned using weak and strong polymeric cation-exchange sorbents. Samples
were washed with 0.5 mL MeOH for 10 seconds. Samples were weighed before and after
each wash. The wash solvent was evaporated to a residual volume of 10 µl and added
50µl 30% methanol.
Samples were air-dried for 24 hours at room temperature after each wash.
Extracts from each wash were analysed by injecting 5 µL into the column
on an Exion UHPLC system that consisted of two Exion AD pumps, an Exion AD
multiplate autosampler set at 10 $\pm$ 2 &deg;C and an Exion AC column oven set
at 40 $\pm$ 2 &deg;C (Sciex, Ontario, Canada).
Separation was performed using a Raptor Biphenyl UHPLC column (2.7 mm, 2.1 mm I.D.
$\times$ 100 mm) (Restek, Bellefonte, PA). The mass spectrometer was a Sciex QTRAP
6500+ with a TurboIonSpray probe for electrospray ionisation.  
The remaining calculus was dissolved using lysing tube beads in 800 $\mu$L of 0.5
$\small{M}$ citric acid (CA) and 50 $\mu$L stable isotope-labelled analogue used
as internal standards (SIL-IS) solution for 1 h at ambient
temperature with gentle shaking. The suspension was then mixed with 800 $\mu$L
MeOH and centrifuged at 10,000 $\times$ g for 5 mins, and analysed by the same
method as the wash extracts.
Data analysis was performed using Analyst 1.7 and MultiQuant 3.0.3 (Sciex).
Raw quantities of compounds are presented in ng and concentrations as ng / mg.

The samples in the replication batch were processed in the same way, but
analysed on equipment used exclusively for oral samples.

@@

The positive correlation between the weight of the calculus
sample and recovered quantities of the compounds suggests sample weight may affect
the ability to detect compounds; although, we were able to detect
compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].

@@

```{r}
#| label: fig-detection-preservation
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    batch == "batch1",
    quant > 0,
    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()
```

```{r}
#| label: fig-detection-preservation2
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    quant > 0,
    batch == "batch2",
    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()

Distribution of state-of-preservation in batches 1 and 2 to make sure the number
of skeletons are not affecting the relationships shown above. Given our sample
contains a smaller number of individuals with fair preservation, this may bias
our interpretations (@fig-preservation-detection).

```{r}
#| label: fig-preservation-detection
uhplc_calculus_long %>% 
  ungroup() %>% 
  distinct(id, batch, .keep_all = TRUE) %>%
  remove_missing(vars = c("preservation", "weight")) %>% 
  ggplot(aes(x = preservation)) +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw()
```

@@

tobacco <- uhplc_calculus_long %>%
  filter(
    #batch == "batch2",
    compound %in% c("nicotine", "cotinine"),
    sex == "m" | sex == "pm"
    ) %>% 
  mutate(detection = if_else(quant == 0, 0, 1)) # 0 = not detected; NA = not included in batch 2.

tobacco_accuracy <- tobacco %>%
  remove_missing(vars = "quant") %>%
  group_by(sample, id, .drop = F) %>% 
  summarise(
    detection = sum(detection),
    #compound = compound
    ) %>% 
  left_join(select(demography, id, pipe_notch, preservation, age), by = "id") %>% 
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(
      detection == 0 & pipe_notch == "N" ~ 1,
      detection > 0 & pipe_notch == "Y" ~ 1,
      detection > 0 & pipe_notch == "N" ~ NaN, # no way of knowing macroscopically if the person smoked without pipe
      TRUE ~ 0
    )
  )

accuracy_age <- tobacco_accuracy %>% # move to supplementary material
  group_by(age) %>% 
  summarise(mean = mean(correct, na.rm = T))

@@

We found no correlation between the number of pipe notches
and the concentration of nicotine and cotinine, suggesting that our ability
to detect tobacco consumption in dental calculus does not necessarily rely on
targeting frequent smokers; here, we consider individuals with multiple pipe notches
as likely to have been heavy smokers.

@@

The presence of pipe notch(es) in an individual and concurrent detection of nicotine
and/or cotinine is used as a crude indicator of the accuracy of the method. When
combining the results of both batches, the method was able to detect some form
of tobacco in
`r nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0))`
of `r nrow(filter(tobacco_accuracy, pipe_notch == "Y"))`
individuals with a pipe notch
(`r scales::percent(nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(tobacco_accuracy, pipe_notch == "Y")), accuracy = 0.1)`).
When also considering correct the absence of a tobacco alkaloid together with the absence
of a pipe notch, the accuracy of the method is
`r scales::percent(mean(tobacco_accuracy$correct, na.rm = T), accuracy = 0.1)`.
Accuracy in the old adult age category is
`r scales::percent(filter(accuracy_age, age == "old")$mean, accuracy = 0.1)`.

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0))`
out of
`r nrow(filter(replicated_accuracy, pipe_notch == "Y"))`
pipe smokers
(`r scales::percent(nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(replicated_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.

@@

Pipe notches were identified by wear on the mesial and distal
sides of the crowns between to teeth, resulting from the practice of clenching a
pipe between adjacent and isomeric teeth, and which differs from the occlusal
wear that occurs through mastication. Wear occurring between adjacent and isomeric
teeth were counted as a single pipe notch.

@@

Caries were scored as the location on each individual tooth. Multiple locations on
a single tooth were separated with `;`. The size of caries was also
recorded, but not used in further analysis. Large caries that cover multiple
surfaces with an unknown origin were recorded as 'crown'. Caries counts were
extracted based on the number of locations recorded.

@@

In the `r caries_ratio_site$n_teeth` that were examined, `r caries_ratio_site$count`
teeth had caries
(`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`).
This frequency has very little meaning, and was further broken down into a ratio
for each individual and each tooth class.

@@

Data on chronic maxillary sinusitis from @casnaUrbanizationRespiratory2021 were
included in this study to assess the relationship between upper respiratory
diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
Chronic maxillary sinusitis (CMS) is the inflammation of the lower paranasal
sinuses, air-filled pockets located in the skull that defend the organism against
inhaled particulate matter and pathogens. This occurs through the production of
mucus carried by small hairs toward an opening situated on the superior part of
the sinus, where pathogens are drained [@slavinDiagnosisManagement2005]. Without
drainage, mucus begins to accumulate in the sinuses, providing an ideal environment
for bacterial growth and thereby contributing to inflammation of the mucous
membranes and subsequently of the bone surfaces [@jangBoneInvolvement2002].
Lesions associated with CMS as defined by @boocockMaxillarySinusitis1995 were
recorded for each individual and classified as "pitting", "spicule-type bone
formation", "remodeled spicules", or "white pitted bone". CMS was scored as absent
when the sinus presented smooth surfaces with little or no associated pitting.
To facilitate inspection, fragmented sinuses were cleaned using a dry tooth-brush
and water where necessary. If the sinuses were not observable with the naked eye,
they were examined with a flexible medical endoscope (Pentax, model: FNL-10RBS,
ø=4mm; view angle=30°) inserted through minor breaks naturally occurring on the
inferior nasal conchae and palatine bone, where the bone tissue is thinner.

<!-- description of other diseases
Osteoarthritis
Vertebral osteophytosis
Cribra orbitalia
Degenerative disc disease
Schmorl's nodes
-->

@@

The sample bias caused by targeting larger calculus deposits may also affect the detection of
compounds, as caffeinated drinks are often
acidic, and may cause a lower calculus formation.
We only found a weak negative correlation
between caffeine concentration and dental calculus index, but since we targeted individuals with
calculus, this may inadvertently have been controlled for during sample collection.

<!-- correlations on caries and calculus separated by quadrant/tooth type -->
